\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-20 paper.tex
Yans updates

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of an 11 sample ARG over a 2.4kb
region of the \textit{Drosophila melanogaster} ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods based on
(A) parsimony, (B) sampling the SMC$\prime$, (C) copying from
inferred ancestral haplotypes, and (D) correlated local tree estimation.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
Position on the X and Y axes for the graph visualization is arbitrary.
Line width and node colour are as described in Fig.~\ref{fig-simplification}.
The graphic below each ARG shows the genomic intervals for each graph edge in the
corresponding ARGs: edges are horizontal lines, stacked in time order
(edges with youngest children at the bottom), with vertical dashed lines indicating
breakpoints between local trees.
}
\end{figure}

The scalability gains made by recent ARG inference methods have been, in part, due to inferring
lower levels of precision about recombination. Indeed, a common statement in the literature is
that modern tools such as \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
construct not an ARG but ``only'' a sequence of local 
trees [TODO cite examples from \#38]. While this statement
is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
adjacent trees to differ, it is certainly not the case that they infer independent local trees.
In fact, both methods take advantage of genome-wide mutation information to construct genealogies,
and as a result nodes and edges persist across multiple local trees.
Their results can therefore be thought of as an inheritance graph, and represented in the gARG format.

To illustrate the varying levels of information that can be captured in a gARG,
Fig.~\ref{fig-inferred-args} presents graph representations of genealogies output by two tools that
\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and 
\argweaver~\citep{rasmussen2014genome}, 
and two which do not (\tsinfer\ and \relate). 
All inferences are based on the
same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
assume a mutation rate of 
% per site per generation may be easier to follow?
\SI{5.49e-9}{site^{-1}generation^{-1}} and a recombination rate of
\SI{8.4e-9}{site^{-1}generation^{-1}} [CITE stdpopsim] and a constant effective population size of 
1,000,000.
The version numbers of inference software, together with the specific inference parameters
needed to reproduce the ARGs, are available from our GitHub repository.

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples 
from the posterior distribution over
ARGs given the data and a fixed model (the SMC$\prime$, an approximation of the full coalescent
with recombination) [TODO CITE]. 
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

>>>>

\begin{figure} \begin{center}
\includegraphics[width=\textwidth]{illustrations/inference.pdf} \end{center}
\caption{\label{fig-inferred-args} Inference of sample-resolved ARGs for 11
\textit{Drosophila melanogaster} genetic sequences over a 2.4kb
region of the ADH locus~\citep{kreitman1983nucleotide}.
Results for four different methods based on
(A) parsimony, (B) sampling the SMC$^\prime$, (C) copying from
inferred ancestral haplotypes, and (D) correlated local tree estimation.
Edge colours are based on the time of the edge's child node
(lighter: older; darker: younger).
X and Y positions of graph nodes are arbitrary;
line width and node colour are as described in Fig.~\ref{fig-simplification}.
Graphics on the bottom row show the genomic positions, relative to the start of the ADH gene,
for each graph edge from the corresponding ARG. Edge intervals are drawn as horizontal lines,
stacked in time order (edges with youngest children at the bottom);
vertical dashed lines indicate breakpoints between local trees.
}
\end{figure}

The scalability gains made by recent ARG inference methods have been, in part, due to inferring
lower levels of precision about recombination. Indeed, a common statement in the literature is
that modern tools such as \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
construct not an ARG but ``only'' a sequence of local 
trees \citep{hejase2020summary, ignatieva2021kwarg}. While this statement
is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
adjacent trees to differ, it is certainly not the case that they infer independent local trees.
In fact, both methods take advantage of genome-wide mutation information to construct genealogies,
and as a result nodes and edges persist across multiple local trees.
Their results can therefore be thought of as an inheritance graph, and represented in the gARG format.

To illustrate the varying levels of information that can be captured in a gARG,
Fig.~\ref{fig-inferred-args} presents graph representations of genealogies output by two tools that
\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and 
\argweaver~\citep{rasmussen2014genome}, 
and two which do not (\tsinfer\ and \relate). 
All inferences are based on the
same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
assume a mutation rate of 
% per site per generation may be easier to follow?
\SI{5.49e-9}{per site per generation} and a recombination rate of
\SI{8.4e-9}{per site per generation} \citep{lauterbur2022expanding}, together with
a constant effective population size of 1,000,000.
The version numbers of inference software, together with the specific inference parameters
needed to reproduce the ARGs, are available from our GitHub repository.

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples 
from the posterior distribution over
ARGs given the data and a fixed model (the SMC$^\prime$, an approximation of the full coalescent
with recombination \citep{marjoram2006fast}).
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).
